Responsive GPU Workgroup Management in the Presence of Persistent Kernels

### 2016-10-13 write-ups/PLDI/paper.tex
Notes from chat today.

@@

>>>>

\ADComment{Points to make: It is important to be able to execute
  long-running, cooperative applications on GPUs (give examples).
  However, resource sharing and resource flexibility is also important
  (give examples, including sharing resources with graphics, sharing
  resources between compute taks, and being flexible about resources
  under energy constraints), and these are not supported by today's
  GPU platforms: long-running applications cause display freezes, etc.
  Resource-based preemption, whereby workgroup execution is postponed
  to free up resources for other purposes (or to reduce resources to
  save energy) can potentially enable resource sharing.  There is
  evidence in the research literature that efficient resource sharing
  for GPUs is possible (cite academic works), and signs from industry
  that this is coming in future architectures.  [Explain more clearly
    what we mean by resource-based preemption.]  Unfortunately,
  arbitrary resource-based preemption is fundamentally incompatible
  with cooperative applications---only trivial examples of cooperation
  are safe [reduction, load balancing via atomic increment]; the HSA
  model of guaranteeing preemption in reverse order of IDs allows some
  cooperative applications to be implemented.  Still, in the general
  case even a mutex cannot be used safely.}

\ADComment{Say that we believe the problem can only be solved through
  programming language innovations with appropriate runtime support.
  We have observed inter-workgroup barriers and work stealing to be
  the main use cases for cooperation, and so in this paper propose two
  new language primitives, ckill and cfork, designed to meet the needs
  of these primitives.  [Describe them a bit.]}

\ADComment{Move on to our main contributions}

\ADComment{This is what we had before.}

@@

\section{Motivating example: work stealing}

\ADComment{Use the work stealing example to both introduce OpenCL and
  also make our key point: that arbitrary preemption will not work.}

\section{The ckill and cfork primitives}

\ADComment{Remind the reader that these are inspired by work stealing
  and barrier synchonization.  In \mysec\ref{ } we describe the
  semantics of kernels that use these primitives.  In \mysec\ref{ } we
  show how work stealing can be implemented.  In \mysec\ref{ } we show
  how a barrier can be implemented.}

\subsection{Semantics of the new primitives}

\ADComment{General semantics: let $N$ be the maximum number of
  workgroups the GPU of interest can support [gloss over or deal with
    issue that this is kernel-dependent].  A kernel is launched with
  $M$ workgroups, for some $0 < M \leq N$.  Initially these are all
  scheduled, and are guaranteed to make independent forward progress.}

\ADComment{The purpose of ckill is for a workgroup to volunteer to
  quit execution of the kernel if the resources that it is using
  (e.g.\ the associated compute unit) are required for other tasks.  A
  ckill statement must be reached under workgroup-uniform control
  flow.  When a workgroup reaches ckill, the workgroup either
  terminates execution (because the environment requires the
  workgroup's resources), or continues execution as if the ckill
  statement were a no=op).  From a semantics perspective this is a
  non-deterministic choice: an application that invokes ckill should
  be prepared for either outcome.  In practice, the effect of ckill
  depends on environmental conditions.}

\ADComment{The purpose of cfork is for a workgroup to indicate that it
  would be useful to have additional workgroups join the computation,
  and to specify that the program point associated with the cfork
  statement would be a suitable place for a new workgroup to start
  executing.  A cfork statement must also be reached under
  workgroup-uniform control flow.  If $C$ workgroups are executing
  when a workgroup reaches a cfork statement, with $0 < C \leq M$,
  then between 0 and $M-C$ new workgroups commence execution from the
  following statement.  From a semantics perspective, the number of
  workgroups that join the computation is nondeterministic, within the
  permitted range.  In practice, how many workgroups join depends on
  environmental conditions.}

\ADComment{TODO: describe persistent variables.}

\ADComment{TODO: explain that ckill and cfork atomically update the
  kernel context (ids).}

\ADComment{TODO: somewhere we need to explain the constraints this
  places on programming: how ids work, what one can assume about them,
  etc.}

\ADComment{Now describe the more refined semantics where IDs are
  relinquished in decreasing order; explain that from a formal
  perspective thie refines the original semantics.}

\ADComment{In the above, explain at some point how a scheduler can
  exploit cfork and ckill to utilise the GPU resources.}

\subsection{Preemption-aware work stealing}

\ADComment{Show how work stealing can now be made safe.}

\subsection{A preemption-aware global barrier}

\ADComment{Show how the global barrier can be written in a manner that
  maintains contiguous ids using the refined semantics.}

\section{Application-specific optimisations}

\ADComment{Explain that for global barriers, we explain how some new
  primitives can lead to a better implementation.  Explain that for
  work stealing there are various ways the existing ckill and cfork
  primitives can be used.}

\subsection{An optimised barrier implementation}

\ADComment{Remark that the barrier of \mysec\ref{sec:naivebarrier} may
  be inefficient, discuss the use of query primitives to make this
  more efficient.}

\subsection{Variations on work stealing}

\ADComment{Three cases here: always ckill/cfork, ckill/cfork only when
  queue is empty, and have a ``mediator'' workgroup to manage things.}

\section{A mega kernel-based prototype}

\ADComment{Explain carefully that doing this properly would require
  driver support, etc., and then explain what we've actually done.}

\section{Experimental evaluation}

\ADComment{TODO: write up notes from whiteboard chat.}

\section{Related work}

\section{Conclusions and future work}

\ADComment{Next-generation GPUs and drivers for evaluation,
  implementation in a real driver, migration of work between multiple
  GPUs.}
